---
title: "Maximizing the potential of high-throughput next-generation sequencing through precise normalization based on read-count distribution"
output: pdf_document
date: "2022-01-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(dplyr)
library(ggpubr)
library(svglite)
library(data.table)
library(tidyverse)
library(cowplot)
library(stringr)
```


```{r}
iseq_norm_counts_arranged <- read.csv("iseq_norm_counts_arranged.csv")
quant_norm_counts_arranged <-  read.csv("quant_norm_counts_arranged.csv")
```


coefficient of variation (CV) of read-count normalization 
```{r}
sd(iseq_norm_counts_arranged$sample_proportion)/mean(iseq_norm_counts_arranged$sample_proportion)
```

coefficient of variation (CV) of fluorescent quantification normalization 
```{r}
sd(quant_norm_counts_arranged$sample_proportion)/mean(quant_norm_counts_arranged$sample_proportion)
```

```{r}
iseq_median_proportion <- median(iseq_norm_counts_arranged$R1_R2_read_proportion)
quant_median_proportion <- median(quant_norm_counts_arranged$R1_R2_read_proportion)
iseq_median_reads <- median(iseq_norm_counts_arranged$R1_R2_reads_combined)
quant_median_reads <- median(quant_norm_counts_arranged$R1_R2_reads_combined)
```

Saving 5% of the number of samples as a variable
```{r}
percent <- (5*350)/100
percent
```


Since the samples are already arranged in order of descending read counts, we can take the top and bottom 5% of samples straight from the data table as the samples with the most and least read counts respectively.
```{r}
top_percent_iseq <- head(iseq_norm_counts_arranged, percent)
top_percent_quant <- head(quant_norm_counts_arranged, percent)
bottom_percent_iseq <- tail(iseq_norm_counts_arranged, percent)
bottom_percent_quant <- tail(quant_norm_counts_arranged, percent)
```


Coupling samples with their replicate from each method
```{r}
matching_quant_samples_iseq_top <- subset(quant_norm_counts_arranged,( Category %in% top_percent_iseq$Category))
matching_quant_samples_iseq_bottom <- subset(quant_norm_counts_arranged,( Category %in% bottom_percent_iseq$Category))
matching_iseq_samples_quant_top <- subset(iseq_norm_counts_arranged,( Category %in% top_percent_quant$Category))
matching_iseq_samples_quant_bottom <- subset(iseq_norm_counts_arranged,( Category %in% bottom_percent_quant$Category))
```


Merge the data from each normalization method
```{r}
merged_data <- merge(iseq_norm_counts_arranged, quant_norm_counts_arranged, by = "Sample")
#head(merged_data)
```

Create scatter plot displaying correlation of reads counts between each method

```{r}
scatter <- ggplot(merged_data, aes(x = sample_proportion.x, y=sample_proportion.y)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  #guides(color=guide_legend(title="Host Subject ID")) +
  xlab("Read Count Norm.
   % of Total Reads") +
  ylab("Fluorescent Quantification Norm.
 % of Total Reads") +
  ggtitle("") +
  scale_x_continuous(breaks = seq(0, 2, by=0.1)) +
  scale_y_continuous(breaks = seq(0, 2, by=0.2)) +
  geom_vline(xintercept=iseq_median_proportion, linetype="dotted", color="red", size = 1) +
  geom_hline(yintercept=quant_median_proportion, linetype="dotted", color="turquoise", size = 1) +
   geom_abline(slope = 1)
scatter
```
#```{r}
save_plot('figure_figure_iseq_abundance_scatter.png',
          scatter,
         base_width = 10,
         base_height = 5)
#```

Bind both datasets from each normalization method

```{r}
top_iseq_matching_quant <- rbind(top_percent_iseq, matching_quant_samples_iseq_top)
bottom_iseq_matching_quant <- rbind(bottom_percent_iseq, matching_quant_samples_iseq_bottom)
top_quant_matching_iseq <- rbind(top_percent_quant, matching_iseq_samples_quant_top)
bottom_quant_matching_iseq <- rbind(bottom_percent_quant, matching_iseq_samples_quant_bottom)
```


Create bar charts displaying samples from each method with most and least reads coupled with their replicates from the other method
```{r}
#Samples from iseq read count norm with most amount of reads
a <- ggplot(top_iseq_matching_quant, aes(x=Sample, y=sample_proportion, fill=method)) +
geom_bar(stat="identity", color="black", position=position_dodge()) +
  theme(axis.text.x = element_blank(),
        legend.position="none",
        axis.ticks.x = element_blank(),
        axis.title.y = element_text(size = 20),
        axis.text.y = element_text(size = 15),
  plot.title = element_text(size = 15))+
  guides(color=guide_legend(title="Top 2% of Samples with most reads from iSeq Norm")) +
  geom_hline(yintercept=c(iseq_median_proportion,quant_median_proportion), linetype=c("dotted","dotted"), color=c("red","turquoise"), size = 1.5) +
  xlab("") +
  ylab("% of total reads") +
  ggtitle("")
a
```

```{r}
#Samples from fluorescent quant norm with most amount of reads
b <- ggplot(top_quant_matching_iseq, aes(x=Sample, y=sample_proportion, fill=method)) +
geom_bar(stat="identity", color="black", position=position_dodge()) +
    theme(axis.text.x = element_blank(),
        legend.position="none",
        axis.ticks.x = element_blank(),
        axis.title.y = element_text(size = 20),
        axis.text.y = element_text(size = 15),
        plot.title = element_text(size = 15)) +
    geom_hline(yintercept=c(iseq_median_proportion,quant_median_proportion), linetype=c("dotted","dotted"), color=c("red","turquoise"), size = 1.5) +
  xlab("") +
  ylab("% of total reads") +
  ggtitle("")
b
```
```{r}
#Samples from fluorescent quant norm with least amount of reads
c <- ggplot(bottom_quant_matching_iseq, aes(x=Sample, y=sample_proportion, fill=method)) +
geom_bar(stat="identity", color="black", position=position_dodge()) +
  theme(axis.text.x = element_blank(),
        legend.position="none",
        axis.ticks.x = element_blank(),
        axis.title.y = element_text(size = 20),
        axis.text.y = element_text(size = 15),
        plot.title = element_text(size = 15)) +
    geom_hline(yintercept=c(iseq_median_proportion,quant_median_proportion), linetype=c("dotted","dotted"), color=c("red","turquoise"), size = 1.5) +
  xlab("") +
  ylab("% of total reads") +
  ggtitle("")
c
```

```{r}
#Samples from iseq read count norm with least amount of reads
d <- ggplot(bottom_iseq_matching_quant, aes(x=Sample, y=sample_proportion, fill=method)) +
geom_bar(stat="identity", color="black", position=position_dodge()) +
  #guides(fill=guide_legend(title="Normalization")) +
  theme(axis.text.x = element_blank(),
        legend.position="none",
        axis.ticks.x = element_blank(),
        axis.title.y = element_text(size = 20),
        axis.text.y = element_text(size = 15),
        plot.title = element_text(size = 15)) +
  scale_x_discrete(breaks = seq(0,1.5, by = 0.05)) +
    geom_hline(yintercept=c(iseq_median_proportion,quant_median_proportion), linetype=c("dotted","dotted"), color=c("red","turquoise"), size = 1.5) +
  xlab("") +
  ylab("% of total reads") +
  ggtitle("")
d
```

Combine plots
```{r}
figure_iseq_abundance_2 <- plot_grid(#a,
                                   b,
                                 #c,
                                 d,
                                 a,
                                 c,
                                 labels = c("A", "C", "B", "D"),
                                 align = "h",
                                 axis = "l",
                                 label_size = 25,
                                 label_fontfamily = 'sans',
                                 ncol = 2,
                                 hjust = -0.25)
figure_iseq_abundance_2
```

#```{r}
save_plot('iseq_abundance.png',
          figure_iseq_abundance_2,
         base_width = 16,
          base_height = 9)
#```


*Creating Rarefaction Curve*

Arranging the data by decreasing read counts
```{r}
iseq_norm_counts_arranged_samples <- iseq_norm_counts_arranged[order(iseq_norm_counts_arranged$R1_R2_reads_combined, decreasing = TRUE),]

quant_norm_counts_arranged_samples <- quant_norm_counts_arranged[order(quant_norm_counts_arranged$R1_R2_reads_combined, decreasing = TRUE),]
head(iseq_norm_counts_arranged_samples)
```


Assigning a rank to these reads counts

```{r}
quant_norm_counts_arranged_samples$rank <- c(1:350)
head(quant_norm_counts_arranged_samples)
iseq_norm_counts_arranged_samples$rank <- c(1:350)
head(iseq_norm_counts_arranged_samples)
```

Combining both data sets

```{r}
total_reads <- rbind(iseq_norm_counts_arranged_samples, quant_norm_counts_arranged_samples)
```


Creating Plot

```{r}
descending_reads_plot <- ggplot(total_reads, aes(x = reorder(rank,-R1_R2_reads_combined), y = R1_R2_reads_combined, group=method)) +
  geom_line(aes(color=method)) +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5)) +
        guides(color=guide_legend(title="")) +
  scale_x_discrete(breaks = seq(0, 400, by = 25)) +
  scale_y_continuous(breaks = seq(0, 60000, by = 10000)) +
  xlab("Number of Samples Retained") +
  ylab("Rarefaction Depth") +
  ggtitle("") +
   geom_hline(yintercept=c(iseq_median_reads,quant_median_reads), linetype="dotted",color = c("red", "turquoise"), size=1)
descending_reads_plot
```

```{r}
rarefaction <- ggplot(total_reads, aes(x = R1_R2_reads_combined, y = rank, group=method)) +
  geom_line(aes(color=method)) +
  theme(axis.text.y = element_text(angle = 0, vjust = 0.5),
        axis.text.x = element_text(angle = 0, vjust = 0.5)) +
        guides(color=guide_legend(title="Normalization")) +
  scale_y_continuous(breaks = seq(0, 400, by = 25)) +
scale_x_continuous(breaks = seq(0, 70000, by = 10000)) +
  ylab("Number of Samples Retained") +
  xlab("Rarefaction Depth") +
  ggtitle("") +
  geom_vline(xintercept=c(iseq_median_reads,quant_median_reads), linetype="dotted",color = c("red", "turquoise"), size=1)
rarefaction
```


#```{r}
save_plot('iseq_rarefaction.png',
          rarefaction,
          base_width = 8,
          base_height = 4)
#```


*Creating boxplot displaying read counts per sample per normalization method*

```{r}
iseq_boxplot <- ggplot(total_reads, aes(x = method, y =sample_proportion)) +
  geom_boxplot(aes(fill=method), outlier.shape=NA) +
  geom_jitter(size=0.8) +
  theme(axis.text.y = element_text(angle = 0, vjust = 0.5),
        axis.text.x = element_text(angle = 0, vjust = 0.5),
        axis.title.x = element_text(size=12),
        axis.title.y = element_text(size=12),
        legend.position="none")  +
        guides(color=guide_legend(title="")) +
  stat_boxplot(geom = "errorbar",
               width = 0.15) + 
  scale_y_continuous(breaks = seq(0, 1.5, by = 0.1)) +
  ylab("% of Total Reads") +
  xlab("Normalization") +
  ggtitle("")
iseq_boxplot
```



#```{r}
save_plot('iseq_boxplot.png',
          iseq_boxplot,
          base_width = 8,
          base_height = 4)
#```
